Investigating soundscapes perception through acoustic scenes simulation.

### 2017-12-18 lafaySimulatedScenePerception.tex
more

\begin{itemize}
\item \emph{experiment 1.a (simulation)}: during this experiment, participants are asked to simulate urban sound environments, using the simulation tool \emph{Simscene}, see Section~\ref{sec:simulator}. Each of them has to create two sound environments: one ideal / pleasant and the other one, non ideal / unpleasant.
\item \emph{experiment 2.a (evaluation)}: after the simulation phase, only a binary information on the pleasantness property of the respective scenes is available: respectively ideal (i) or non ideal (ni). Furthermore, this information is given by the creator of the scene. The second experimental step aims at investigating more deeply and more broadly our knowledge on the pleasantness of the simulated scenes. For that, a second group of participants is asked to evaluate the pleasantness of each scene, on a semantic scale. This experiment has two goals:
\begin{enumerate}
\item to evaluate more precisely the respective influence of the various sources composing the scenes on the pleasantness  (i or ni) thanks to a finer quantification of the pleasantness of the scene.
\item to detect the presence of \emph{outliers} or ambiguous scenes. Indeed, throughout our analyses, the predefined hedonic properties of the scenes (i and ni) are used as reference. We thus need to ensure beforehand that no ambiguity exists between extreme cases of i and ni scenes, \ie~that the less pleasant i-scene remains above the more pleasant ni-scene.
\end{enumerate}
\end{itemize}

@@

The participants have to simulate two urban sound environments of one minute each. They follow the instructions below:

@@

All the participants start by designing the ideal environment; they read the second set of instructions at the end of the first experiment. Participants are completely free of their choices concerning sounds and synthesis parameters. The created sound environments must nevertheless fulfill the two following constraints :

@@

\begin{itemize}
\item the listening point of view is the one of a fixed listener;
\item the soundscape must be realistic, \ie~physically plausible. For instance, participants are free to insert ten dogs in the soundscape but they cannot insert one dog barking every 10 milliseconds.
\end{itemize}

@@

In addition, once the two sound scenes are completed, participants are invited to:

@@

Before starting the first simulation, a 20-minute tutorial is given to the participants in order to familiarize them with the simulation interface and the sound database. The experiment is planned to last about two hours and a half, including breaks that the subjects are allowed to take.

@@

All the participants perform the experiment on standard desktop computers with the same hardware and software configurations. The audio files are played in diotic conditions using headphones. During the tutorial, participants are asked to adjust the sound volume to a comfortable level. Once set, they are not allowed to modify it during the remaining of the experiment.

@@

All the participants performed the experiment at the same time. They are equally distributed in three identical quiet rooms. They are not allowed to talk to each other during the experiment.

@@

Three experimenters (one in each room) are available during the whole duration of the experiment in order to assist participants with potential hardware and software issues and to answer questions.

\subsubsection*{Participants}

@@

Among the 44 participants, 40 succeeded, producing at the end 80 simulated sound scenes (40 ideal scenes, 40 non ideal scenes). the 4 other participants were excluded from the process due to a lack of understanding of the instructions or failure to respect them, or for exceeding the maximum duration allowed to perform the experiment.

@@

Due to temporal constraints, participants only assess 30 seconds of the initial 1-minute simulated scenes generated in experiment 1.a (from sec. 15 to sec. 45).

@@

The assessment is done with a 7-point bipolar semantic scale going from -3 (non ideal / unpleasant) to +3 (ideal / pleasant). Before evaluating a scene, participants must listen to the first 20 seconds of the stimuli. After the evaluation, they are free to continue to the next scene.

@@

For each participant, sound scenes are played in a quasi random order. 5 ideal scenes and 5 non ideal scenes are first sequenced to allow the participants to calibrate their scores. These first 10 scenes are played back again at the end of the experiment. Only the last evaluations are taken into account.

@@

The experiment is planned to last 30 minutes. The participants do not know anything beforehand about the nature of the sound scenes.

@@

All the participants perform the experiment on computers with equivalent hardware and software configurations. The audio files are played in diotic conditions by semi-open headphones \emph{Beyerdynamic DT 990 Pro}. The stimuli are the scenes obtained by the simulation experiment of experiment 1.a. The output sound level is the same for all the participants.

@@

All the participants perform the experiment simultaneously. They are not allowed to talk to each other during the experiment.

@@

An experimenter is available during the whole duration of the experiment in order to assist participants and answer questions any questions the participants may have.

\subsubsection*{Participants}

@@

All the participants succeeded in doing the experiment.

@@

\begin{itemize}
\item \emph{perceptual features}: the perceived pleasantness of the composed scene, assessed on a 7-point bipolar semantic scale. $\mathcal{A}_{scene}$ is the average pleasantness for one scene, as the average of the marks of all the participants. $\mathcal{A}_{subject}$ is the average pleasantness for each participant, computed as the average of all the score given by each participant. Given the low number of participants in experiment 1.b, we choose not to normalize the pleasantness scores.

@@

First, in order to ensure the coherence of the data, we check that none of the ni-scenes gets a $\mathcal{A}_{scene}$ higher than one of a i-scene. 4 ni-scenes don't fulfill that constraint: they are thus removed, together with their corresponding i-scenes. As a consequence, 36 i-scenes and 36 ni-scenes remain for analysis. Second, we verify that participants really perceived a difference in terms of pleasantness between i- and ni-scenes. For that, we investigate the mean pleasantness score for each participant $\mathcal{A}_{subject}$, computed separately for each type of environment. It indeed appears that the i-scenes were perceived significantly more pleasant ($p < 0.01$) than the ni-scenes.

@@

It appears that the biggest influence on the global sound levels comes from the events, the difference between $L$ and $L(E)$ being only 1 $dB$ for i- and ni-scenes. This observation is in agreement with the results obtained by Kuwano~\al~\cite{kuwano_memory_2003}. During their experiment, the authors ask their participants to assess a set of soundscapes at a global level and then to do the same judgment at the time when they detect a sound source. The study shows that there is no significant difference between global and averaged instantaneous judgments. In our case, the result can be interpreted as if the participants had unconsciously integrated this perceptual reality when composing the scenes, by allocating most of the global sound levels to well identified and relatively short sounds, \ie the events.

@@

We study the composition of the scenes by counting the number of participants who used a given class of sounds to simulate the environment. Results are shown on Figure~\ref{fig:soundsourcea} for events and on Figure~\ref{fig:soundsourceb} for textures. For the sake of clarity, a transitional level of abstraction between level 0 and 1, named $0+$ is used to depict classes, see ~Figures~\ref{fig:taxonomieEa}, \ref{fig:taxonomieEb} and \ref{fig:taxonomieT}.

@@

We observe a significant difference between i- and ni-scenes for the classes choice. The distribution of the classes is very similar to the one obtained in a related work on ideal urban soundscapes \cite{guastavino2006ideal}, \ie~classes involving human presence and nature are prevailing in the i-scenes, and on the opposite, classes involving mechanical sounds and/or public works are prevailing in the ni-scenes. These results confirm a previously observed fact: the semantic nature of the sound sources play an important role in the assessment of the environment \cite{raimbault2005urban,dubois2006cognitive}.

@@

To a certain extent, our results are in contradiction with this assumption. In fact,  Figure~\ref{fig:soundsourcea} show that \emph{public transportation} classes (\emph{bus} and \emph{train}, \cf~Figure~\ref{fig:soundsourcec}) have been used by the participants for $28\%$ of the i-scenes, and for $42\%$ of the ni-scenes. the results do not re-evaluate the fact that sounds of \emph{public transportation} are well accepted: $25\%$ of the subjects used \emph{bus} for the i-scenes, a level that is comparable to the \emph{bike} class, and much higher than all the personal vehicles classes. Though, \emph{public transport} classes are also strongly present in the ni-scenes, for instance more than \emph{light vehicle} or \emph{trucks} classes. On the basis of our results, the public transportation class cannot be considered as typical of an ideal urban soundscape.

@@

This difference may be explained by the nature of the experimental protocol used in the two studies. As in our study, Guastavino asks the participants for describing an environment starting from their memories whereas, in our case, participants perform the same task using actual sound samples. The fact that participants in our experiment are faced to the acoustic reality of the sounds for composing the environment may have decreased the socio-cultural impact. Other works using sounds as stimuli have shown that the \emph{bus} class can have a negative influence on the assessment of the environment \cite{lavandier2006contribution}.

@@

\begin{table}[t]
 \setlength{\tabcolsep}{0.2pt}
 \centering
  {\renewcommand{\arraystretch}{0.9}
\begin{tabular}{c c c c}
Abstraction        & \multicolumn{2}{c}{Event sound markers} \\
level & i-scenes & ni-scenes \\
\hline
0  &                          & construction work (3.78)  \\
\hline
  & church bell  (4.5)             & car horn  (3.9) \\
1 & bicycle bell  (4.3)      & siren (3.9)\\
  & animal (4.2)              &       \\
   \hline
  & bird        (4.8)       & car horn  (4.0)\\
2 & church bell  (4.4)             & siren (4.0)\\
  & bicycle bell     (4.2)             &       \\
   \hline
  & bird song (4.8)        & klaxon  (4.1)\\
3 & church bell   (4.3)            & siren (4.0)\\
  & bicycle bell      (4.2)   &       \\
  & foot steps  (3.6)      &  \\
  &                           & \\
  & \multicolumn{2}{c}{Texture sound markers}      \\
  & i-scenes & ni-scenes \\
\hline
0 &     courtyard / park (4.1) &  construction (3.9)  \\
\hline
1 &     park (3.65)          &  crossroad (3.6)  \\
  &                          &  construction vehicule (3.3)  \\
\hline
2 &     park (3.64)          &  crossroad (3.56)  \\
\hline
\end{tabular}
}
\vspace{0.5mm}
\caption{Event and texture classes identified as sound markers. In each cell, markers are ranked in the decreasing order of V-test value. $p\leq0.01$ for all sound markers.}
\label{tab:markers}
\end{table}

@@

We highlight that, on a qualitative point of view, the composition of the scenes in terms of sound sources differs with respect to the type of environment (i or ni). Let us now investigate whether some of the sound classes are specific to a given environment. For that, we use the V-test as detailed in  Section~\ref{sec:xp1_dataAna} separately for each abstraction level. Results are presented in Table~\ref{tab:markers}.

@@

Regarding the sound events, 9 markers are identified for all abstraction levels. As shown on Figure~\ref{fig:soundsource}, classes related to human presence (\emph{male footsteps on concrete}, \emph{bicycle bell}), and of nature (\emph{animals, bird}, and \emph{bird song}) are i-scenes markers as well as the \emph{church bell} class. This latter result may be due to the socio-cultural background of the participants who are mostly European citizens. In fact, according to Schafer, a sound that is identified by a person as being an element of his/her environment, is well accepted. Sound markers of ni-scenes are classes related to construction site (\emph{construction works}), or suggesting a strong traffic (\emph{horn}, \emph{siren}).

@@

Regarding the textures, 5 markers are identified. For the i-scenes, they are classes related to passive or quiet ambiances (\emph{courtyard}, \emph{park}). For the ni-scenes, as for the events, they are classes related to construction site (\emph{construction}, \emph{construction vehicle}), together with a class related to traffic (\emph{crossroad}).

@@

Although the whole set of identified markers are rather intuitive, none of the event classes related to the noise of the motor vehicles are identified as markers, except for the texture class \emph{crossroad}. To simulate an unpleasant traffic, participants chose the classes \emph{horn} or \emph{siren}. We thus conclude that isolated motor vehicle sounds are understood as being part of the urban environment, and thus their nature is not necessarily linked to an unpleasant soundscape.

@@

In this section, we evaluate the ability of a semantic representation of the scenes to discriminate between the two types of environment. For that, we consider a rank 5-precision ($p@5$) on the space induced by the semantic features $S$, and for each abstraction level, see Section~\ref{sec:xp1_dataAna}. The vectors $S$ are built by using all the classes ($ET$), only the event classes ($E$), only the texture classes ($T$), only the event classes corresponding to sound markers ($E_m$), or only the event classes excluding sound markers ($E_{w/o,m}$). Texture classes corresponding to sound markers are not considered as these latter are too few to reliably compute the metric. For the same reasons, event classes corresponding to sound markers at abstraction level $0$ are also discarded. Results are shown on Figure~\ref{fig:pa5}.

@@

Concerning $ET$, the $p@5$ is $76\%$ for the abstraction level $0$, and stay above $86\%$ for the abstraction level $1$. These results confirm that it is possible to make a clear distinction between the two types of environment on the sole basis of presence / absence of sound classes. We also notice that the higher the abstraction level is, the more important the ability to make this distinction is. In other words, the more precise the scene description is, the more clear the discrimination between i- and ni-scenes is.

@@

\subsubsection*{Participants}

@@

%\nm{a ce propos, il n'est pas fait mention des sons manquants pour les participants (cf. protocole expe de 1a);
%ne faudrait -il pas en dire 2 mots ce qui alimenterait aussi la discussion sur la dependance de la qualite (et l'efficacite) de la simulation a  la constitution de la base de donnees de textures et d'evenemetns ?}

>>>>

\begin{itemize}
\item \emph{experiment 1.a (simulation)}: during this experiment, subjects are asked to simulate urban sound environments, using the simulation tool \emph{Simscene}, see Section~\ref{sec:simulator}. Each of them has to create two sound environments: one ideal / pleasant and the other one, non ideal / unpleasant.
\item \emph{experiment 2.a (evaluation)}: after the simulation phase, only a binary information on the pleasantness property of the respective scenes is available: respectively ideal (i) or non ideal (ni). Furthermore, this information is given by the creator of the scene. The second experimental step aims at investigating more deeply and more broadly our knowledge on the pleasantness of the simulated scenes. For that, a second group of subjects is asked to evaluate the pleasantness of each scene, on a semantic scale. This experiment has two goals:
\begin{enumerate}
\item to evaluate more precisely the respective influence of the various sources composing the scenes on the pleasantness  (i or ni) thanks to a finer quantification of the pleasantness of the scene.
\item to detect the presence of \emph{outliers} or ambiguous scenes. Indeed, throughout our analyses, the predefined hedonic properties of the scenes (i and ni) are used as reference. We thus need to ensure beforehand that no ambiguity exists between extreme cases of i and ni scenes, \ie~that the less pleasant i-scene remains above the more pleasant ni-scene.
\end{enumerate}
\end{itemize}

@@

The subjects have to simulate two urban sound environments of one minute each. They follow the instructions below:

@@

All the subjects start by designing the ideal environment; they read the second set of instructions at the end of the first experiment. subjects are completely free of their choices concerning sounds and synthesis parameters. The created sound environments must nevertheless fulfill the two following constraints :

@@

\begin{itemize}
\item the listening point of view is the one of a fixed listener;
\item the soundscape must be realistic, \ie~physically plausible. For instance, subjects are free to insert ten dogs in the soundscape but they cannot insert one dog barking every 10 milliseconds.
\end{itemize}

@@

In addition, once the two sound scenes are completed, subjects are invited to:

@@

Before starting the first simulation, a 20-minute tutorial is given to the subjects in order to familiarize them with the simulation interface and the sound database. The experiment is planned to last about two hours and a half, including breaks that the subjects are allowed to take.

@@

All the subjects perform the experiment on standard desktop computers with the same hardware and software configurations. The audio files are played in diotic conditions using headphones. During the tutorial, subjects are asked to adjust the sound volume to a comfortable level. Once set, they are not allowed to modify it during the remaining of the experiment.

@@

All the subjects performed the experiment at the same time. They are equally distributed in three identical quiet rooms. They are not allowed to talk to each other during the experiment.

@@

Three experimenters (one in each room) are available during the whole duration of the experiment in order to assist subjects with potential hardware and software issues and to answer questions.

\subsubsection*{subjects}

@@

Among the 44 subjects, 40 succeeded, producing at the end 80 simulated sound scenes (40 ideal scenes, 40 non ideal scenes). the 4 other subjects were excluded from the process due to a lack of understanding of the instructions or failure to respect them, or for exceeding the maximum duration allowed to perform the experiment.

@@

Due to temporal constraints, subjects only assess 30 seconds of the initial 1-minute simulated scenes generated in experiment 1.a (from sec. 15 to sec. 45).

@@

The assessment is done with a 7-point bipolar semantic scale going from -3 (non ideal / unpleasant) to +3 (ideal / pleasant). Before evaluating a scene, subjects must listen to the first 20 seconds of the stimuli. After the evaluation, they are free to continue to the next scene.

@@

For each participant, sound scenes are played in a quasi random order. 5 ideal scenes and 5 non ideal scenes are first sequenced to allow the subjects to calibrate their scores. These first 10 scenes are played back again at the end of the experiment. Only the last evaluations are taken into account.

@@

The experiment is planned to last 30 minutes. The subjects do not know anything beforehand about the nature of the sound scenes.

@@

All the subjects perform the experiment on computers with equivalent hardware and software configurations. The audio files are played in diotic conditions by semi-open headphones \emph{Beyerdynamic DT 990 Pro}. The stimuli are the scenes obtained by the simulation experiment of experiment 1.a. The output sound level is the same for all the subjects.

@@

All the subjects perform the experiment simultaneously. They are not allowed to talk to each other during the experiment.

@@

An experimenter is available during the whole duration of the experiment in order to assist subjects and answer questions any questions the subjects may have.

\subsubsection*{subjects}

@@

All the subjects succeeded in doing the experiment.

@@

\begin{itemize}
\item \emph{perceptual features}: the perceived pleasantness of the composed scene, assessed on a 7-point bipolar semantic scale. $\mathcal{A}_{scene}$ is the average pleasantness for one scene, as the average of the marks of all the subjects. $\mathcal{A}_{subject}$ is the average pleasantness for each participant, computed as the average of all the score given by each participant. Given the low number of subjects in experiment 1.b, we choose not to normalize the pleasantness scores.

@@

First, in order to ensure the coherence of the data, we check that none of the ni-scenes gets a $\mathcal{A}_{scene}$ higher than one of a i-scene. 4 ni-scenes don't fulfill that constraint: they are thus removed, together with their corresponding i-scenes. As a consequence, 36 i-scenes and 36 ni-scenes remain for analysis. Second, we verify that subjects really perceived a difference in terms of pleasantness between i- and ni-scenes. For that, we investigate the mean pleasantness score for each participant $\mathcal{A}_{subject}$, computed separately for each type of environment. It indeed appears that the i-scenes were perceived significantly more pleasant ($p < 0.01$) than the ni-scenes.

@@

It appears that the biggest influence on the global sound levels comes from the events, the difference between $L$ and $L(E)$ being only 1 $dB$ for i- and ni-scenes. This observation is in agreement with the results obtained by Kuwano~\al~\cite{kuwano_memory_2003}. During their experiment, the authors ask their subjects to assess a set of soundscapes at a global level and then to do the same judgment at the time when they detect a sound source. The study shows that there is no significant difference between global and averaged instantaneous judgments. In our case, the result can be interpreted as if the subjects had unconsciously integrated this perceptual reality when composing the scenes, by allocating most of the global sound levels to well identified and relatively short sounds, \ie the events.

@@

We study the composition of the scenes by counting the number of subjects who used a given class of sounds to simulate the environment. Results are shown on Figure~\ref{fig:soundsourcea} for events and on Figure~\ref{fig:soundsourceb} for textures. For the sake of clarity, a transitional level of abstraction between level 0 and 1, named $0+$ is used to depict classes, see ~Figures~\ref{fig:taxonomieEa}, \ref{fig:taxonomieEb} and \ref{fig:taxonomieT}.

@@

We observe a noticable difference in terms of class choices between the i- and ni-scenes. The distribution of the classes is very similar to the one obtained in a related work on ideal urban soundscapes \cite{guastavino2006ideal}, \ie~ on one hand, classes involving human presence and nature are prevailing in the i-scenes, and on the other hand, classes involving mechanical sounds and/or public works are prevailing in the ni-scenes. These results confirm a previously observed fact: the semantic nature of the sound sources play an important role in the assessment of the environment \cite{raimbault2005urban,dubois2006cognitive}.

@@

To a certain extent, our results contradict this result. In fact,  Figure~\ref{fig:soundsourcea} show that \emph{public transportation} classes (\emph{bus} and \emph{train}, \cf~Figure~\ref{fig:soundsourcec}) have been used by the subjects for $28\%$ of the i-scenes, and for $42\%$ of the ni-scenes. Those results do not state the fact that sounds of \emph{public transportation} are well accepted: $25\%$ of the subjects used \emph{bus} for the i-scenes, a level that is comparable to the \emph{bike} class, and much higher than all the personal vehicles classes. Though, \emph{public transport} classes are also strongly present in the ni-scenes, for instance more than \emph{light vehicle} or \emph{trucks} classes. On the basis of our results, the public transportation class cannot be considered as typical of an ideal urban soundscape.

@@

This difference may be explained by the nature of the experimental protocol used in the two studies. As in our study, Guastavino asks the subjects to describe an environment. She asks them to perform this task starting only from their memories whereas, in our case, subjects perform the same task using actual sound samples that they can listen to. The fact that subjects in our experiment are faced to the acoustic reality of the sounds for composing the environment may have decreased the socio-cultural impact. Other studies that considered sounds as stimuli have shown that the \emph{bus} class can have a negative influence on the assessment of the environment \cite{lavandier2006contribution}.

@@

\begin{table}[t]
 \setlength{\tabcolsep}{0.2pt}
 \centering
  {\renewcommand{\arraystretch}{0.9}
\begin{tabular}{c c c c}
Abstraction        & \multicolumn{2}{c}{Event sound markers} \\
level & i-scenes & ni-scenes \\
\hline
0  &                          & construction work (3.78)  \\
\hline
  & church bell  (4.5)             & car horn  (3.9) \\
1 & bicycle bell  (4.3)      & siren (3.9)\\
  & animal (4.2)              &       \\
   \hline
  & bird        (4.8)       & car horn  (4.0)\\
2 & church bell  (4.4)             & siren (4.0)\\
  & bicycle bell     (4.2)             &       \\
   \hline
  & bird song (4.8)        & klaxon  (4.1)\\
3 & church bell   (4.3)            & siren (4.0)\\
  & bicycle bell      (4.2)   &       \\
  & foot steps  (3.6)      &  \\
  &                           & \\
  & \multicolumn{2}{c}{Texture sound markers}      \\
  & i-scenes & ni-scenes \\
\hline
0 &     courtyard / park (4.1) &  construction (3.9)  \\
\hline
1 &     park (3.65)          &  crossroad (3.6)  \\
  &                          &  construction vehicule (3.3)  \\
\hline
2 &     park (3.64)          &  crossroad (3.56)  \\
\hline
\end{tabular}
}
\vspace{0.5mm}
\caption{Event and texture classes identified as sound markers. In each cell, markers are ranked as decreasing order of V-test value. $p\leq0.01$ for all sound markers.}
\label{tab:markers}
\end{table}

@@

We have shown that, on a qualitative point of view, the composition of the scenes in terms of sound sources differs i- or ni-scenes. We now investigate whether some of the sound classes are specific to a given environment. For that purpose, the V-test detailed in  Section~\ref{sec:xp1_dataAna} is considered separately for each abstraction level. Results are presented in Table~\ref{tab:markers}.

@@

Regarding the sound events, 9 markers are identified for all abstraction levels. As shown on Figure~\ref{fig:soundsource}, classes related to human presence (\emph{male footsteps on concrete}, \emph{bicycle bell}), and of nature (\emph{animals, bird}, and \emph{bird song}) are i-scenes markers as well as the \emph{church bell} class. This latter result may be due to the socio-cultural background of the subjects who are mostly European citizens. In fact, according to Schafer, a sound that is identified by a person as being an important element of his/her environment, is well accepted. Sound markers of ni-scenes are classes related to construction site (\emph{construction works}), or suggesting a strong traffic (\emph{horn}, \emph{siren}).

@@

Regarding the textures, 5 markers are identified. For the i-scenes, those classes related to passive or quiet ambiances (\emph{courtyard}, \emph{park}). The marker classes for the ni-scenes are, as for the events, related to construction site (\emph{construction}, \emph{construction vehicle}), together with a class related to traffic (\emph{crossroad}).

@@

Although the whole set of identified markers are rather intuitive, none of the event classes related to the noise of the motor vehicles are identified as markers, except for the texture class \emph{crossroad}. To generate an unpleasant traffic, subjects chose the classes \emph{horn} or \emph{siren}. We thus conclude that isolated motor vehicle sounds are understood as being part of the urban environment, and thus their nature is not necessarily linked to an unpleasant soundscape.

@@

In this section, we evaluate at which level a semantic representation of the scenes allows us to discriminate between the two types of environment. For this purpose, a rank 5-precision ($p@5$) is computed on the space induced by the semantic features $S$, and for each abstraction level, see Section~\ref{sec:xp1_dataAna}. The vectors $S$ are built by using all the classes ($ET$), only the event classes ($E$), only the texture classes ($T$), only the event classes corresponding to sound markers ($E_m$), or only the event classes excluding sound markers ($E_{w/o,m}$). Texture classes corresponding to sound markers are not numerous enough to reliably compute the metric. They are thus not considered. For the same reasons, event classes corresponding to sound markers at abstraction level $0$ are also discarded. Results are shown on Figure~\ref{fig:pa5}.

@@

Concerning $ET$, the $p@5$ is $76\%$ for the abstraction level $0$, and remains above $86\%$ for higher abstraction levels. Considering only the presence / absence of sound classes thus allows us to nicely discriminate between the two types of environments. We also notice that the higher the abstraction level is, thus the more precise the description, the more efective the description is.

@@

\subsubsection*{subjects}

@@

%\nm{a ce propos, il n'est pas fait mention des sons manquants pour les subjects (cf. protocole expe de 1a);
%ne faudrait -il pas en dire 2 mots ce qui alimenterait aussi la discussion sur la dependance de la qualite (et l'efficacite) de la simulation a  la constitution de la base de donnees de textures et d'evenemetns ?}
